---
title: "Fungal sensitivity to fungicides in R"
description: "How to estimate EC50 in R"
author: "Kaique Alves"
date: "2022-05-08"
categories:
  - English
  - plant pathology
  - code
  - analysis
---

Hello everyone!

In this post I will show how to estimate fungicide $EC_{50}$ values in `R`. In dose-response terms, $EC_{50}$ is the concentration at which a compound inhibits 50% of growth relative to the untreated control.

In plant pathology, $EC_{50}$ is widely used to evaluate fungal sensitivity to fungicides. An increase in $EC_{50}$ for part of a population may indicate selection toward less sensitive isolates.

# Load the packages

The core model-fitting engine used here is `drc`, which provides several dose-response models. For stratified, multi-isolate workflows, I also recommend the [`ec50estimator`](https://alvesks.github.io/ec50estimator/){target="_blank"} project, which was designed specifically to streamline $EC_{50}$ estimation in plant pathology datasets. If you do not have the packages installed yet, run `install.packages("drc")` in your console and install the others as needed.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(drc)
library(ec50estimator)
library(cowplot)
library(ggridges)
library(ggthemes)
```

# Data

First, we need a dataset from which $EC_{50}$ can be estimated. Here I will simulate data using a standard fungicide dose-response relationship (Noel et al. 2018). If you already have your own dataset, you can import it directly in `R` and move to the modeling steps.

$$ f(x) = c +\frac{d-c}{1+exp(b(log(x)-log(e)))}$$

This is the four-parameter log-logistic model, `LL.4()`. Its parameters are `b`, `c`, `d`, and `e`. Parameter `b` is the slope around the inflection point, `c` is the lower asymptote, `d` is the upper asymptote, and `e` corresponds to the $EC_{50}$. Variable `x` represents the fungicide dose or concentration.

Using this model, we can create a helper function to simulate mycelial growth values.

```{r}
grow = function(dose, b,c,d,e){
  x = dose
  y = c + (d-c)/(1+exp(b*(log(x)-log(e))))
  erro = (y * 0.02 * ( (d + 0.05 * d) - y ))
 
  
  growth = rnorm(length(x),y, erro)
  return(growth)
}
```

Below is a simple example. I create a vector of doses called `doses`, simulate growth values, and inspect the resulting data frame.

```{r}
# create dose vector
doses = rep(c(0, 1e-6,1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1),5)
set.seed(1) # keeps the simulated values reproducible

# simulate mycelial growth
growth = grow(doses, b = 1, c = 0, d = 20, e = 0.001)

df1 = data.frame(doses,growth)
head(df1)
```

We can then visualize the simulated observations.

```{r}
df1 %>% 
  ggplot()+
  geom_jitter(aes(factor(doses),growth, color = growth),
              width = 0.05,
              size = 3.5, 
              alpha = 1)+
  geom_boxplot(aes(factor(doses),growth), fill = NA, size = .7)+
  scale_color_viridis_c(option = "D",direction = 1)+
  labs(x ="Dose", y = "Mycelial growth")+
  theme_minimal_hgrid(font_size = 12)+
  theme(legend.position = "none")
```

# Estimating $EC_{50}$ for a single isolate

To estimate $EC_{50}$, we first fit an appropriate dose-response model. Here we use `drm()` from the `drc` package. We provide:

- the formula, which maps response and dose columns
- the data frame
- the model function used for fitting

In this example, the formula is `growth ~ doses`, the data frame is `df1`, and the model is `LL.4()`.

```{r}
model1 = drm(formula = growth~doses,  data = df1, fct = LL.4())
summary(model1)
```

The fitted curve can be inspected visually.

```{r}
plot(model1)
```

To obtain the $EC_{50}$ estimate itself, we use the function `ED()`. The argument `respLev = 50` requests the effective dose corresponding to 50% response reduction. Confidence intervals can also be obtained with `interval = "delta"`. This low-level workflow is useful because it makes the underlying estimation step explicit before moving to larger automated pipelines such as [`ec50estimator`](https://alvesks.github.io/ec50estimator/){target="_blank"}.

```{r}
ED(model1, respLev=50, interval = "delta")
```

# Estimating $EC_{50}$ for multiple isolates

In practice, we often need to estimate $EC_{50}$ for many isolates rather than a single one. To illustrate this, I will create a hypothetical dataset with 50 fungal isolates, half collected in an organic field and half in a conventional field. These isolates will be evaluated against two fungicides, here called `Fungicide A` and `Fungicide B`.

## Fungicide A

For `Fungicide A`, I assume isolates from the conventional field are less sensitive.

```{r}
set.seed(1)
n_isolates = 25

nrep = 5
n_dose =7
n_field = 2
n = n_isolates*nrep*n_dose*n_field
field = c(rep("Organic",nrep*n_dose), rep("Conventional",nrep*n_dose))
isolate = sort(rep(1:((n_isolates*2)),nrep*n_dose))
doses = (rep(c(0,1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1),nrep*n_field))
EC50 = c(5e-3, 1e-1)


fungicide_A = data.frame(isolate,
           field = rep(field,n_isolates),
           fungicide = "Fungicide A",
           dose = rep(doses,n_isolates)) %>% 
  mutate(effect = case_when(
    field == "Conventional" ~ rnorm(n,EC50[2],0.0001),
    field == "Organic" ~ rnorm(n,EC50[1],0.0001))) %>% 
  mutate(growth = grow(dose, b = 1, c = 0, d = 20, e = effect))%>% 
  dplyr::select(-effect)

head(fungicide_A)
```

## Fungicide B

For `Fungicide B`, I assume both field types have the same sensitivity pattern.

```{r}
set.seed(1)
n_isolates = 25
nrep = 5
n_dose =7
n_field = 2
n = n_isolates*nrep*n_dose*n_field
field = c(rep("Organic",nrep*n_dose), rep("Conventional",nrep*n_dose))
isolate = sort(rep(1:((n_isolates*2)),nrep*n_dose))
doses = (rep(c(0,1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1),nrep*n_field))
EC50 = c(5e-3, 5e-3)


fungicide_B = data.frame(isolate,
                        field = rep(field,n_isolates),
                        fungicide = "Fungicide B",
                        dose = rep(doses,n_isolates)) %>% 
  mutate(effect = case_when(
    field == "Conventional" ~ rnorm(n,EC50[2],0.0001),
    field == "Organic" ~ rnorm(n,EC50[1],0.0001))) %>% 
  mutate(growth = grow(dose, b = 1, c = 0, d = 20, e = effect)) %>% 
  dplyr::select(-effect)

head(fungicide_B)
```

Now we combine the two data frames.

```{r message=FALSE, warning=FALSE}
fungicide = fungicide_A %>% 
  bind_rows(fungicide_B)
head(fungicide)
```

There is a way to fit all isolates at once using `drm()` with `curveid = isolate`. However, with large datasets this may become slow, and the resulting object can be difficult to handle.

```{r eval=FALSE, include=TRUE}
# DO NOT RUN THIS ON VERY LARGE DATASETS WITHOUT CHECKING PERFORMANCE
model = drm(growth~dose, curveid = isolate,  fct = LL.4(), data = fungicide)
ED(model, 50, interval = "delta")
```

For more practical workflows, I prefer using [`ec50estimator`](https://alvesks.github.io/ec50estimator/){target="_blank"} rather than writing a loop around `drm()` manually. That is exactly the use case the package was designed for: estimating $EC_{50}$ isolate by isolate within structured datasets while preserving grouping variables and confidence intervals in a tidy output.

In other words, `drc` is excellent for understanding and fitting an individual curve, whereas `ec50estimator` becomes the practical layer when the analysis involves many isolates, fungicides, or production systems.

```{r}
EC50_est <- estimate_EC50(
  growth ~ dose,
  data = fungicide,
  isolate_col = "isolate",
  strata_col = c("field", "fungicide"),
  interval = "delta",
  fct = drc::LL.4()
) %>% 
  as.data.frame()

head(EC50_est)
```

Important note: in this post I used the same model, `LL.4()`, for both fungicides. In real applications, you should verify which model is most appropriate for your data. A useful strategy is to fit candidate models with `drm()` for a subset of isolates and compare them with `mselect()`. Once that modeling choice is clear, `ec50estimator` makes it much easier to apply the workflow consistently across all curves.

## Visualizing the estimated EC50 values

Once estimates are available, one useful step is to compare their distributions across strata.

```{r fig.width=8, fig.height=4.8}
EC50_est %>% 
  ggplot(aes(x = field, y = Estimate, color = field)) +
  geom_jitter(width = 0.1, alpha = 0.6) +
  geom_boxplot(fill = NA, outlier.shape = NA) +
  facet_wrap(~fungicide, scales = "free_x", nrow = 1) +
  theme_half_open(font_size = 12) +
  scale_color_colorblind() +
  scale_y_log10() +
  labs(
    x = "Production system",
    y = expression(EC[50]),
    title = expression("Estimated distributions of " * EC[50] * " across production systems")
  )
```

Another possibility is to inspect the full distribution more carefully.

```{r fig.width=8, fig.height=5.6}
EC50_est %>% 
  ggplot(aes(x = Estimate, y = field, fill = field)) +
  geom_density_ridges(alpha = 0.7) +
  facet_wrap(~fungicide, nrow = 2) +
  theme_half_open(font_size = 12) +
  scale_fill_colourblind() +
  scale_x_log10() +
  labs(
    x = expression(EC[50]),
    y = "Production system",
    title = expression("Density view of estimated " * EC[50] * " values")
  )
```

# Final remarks

Estimating $EC_{50}$ is an important step in fungicide sensitivity studies, but good inference depends on more than simply fitting one model and reporting one number. It is important to think about:

- the biological meaning of the response variable
- the appropriateness of the selected dose-response model
- the structure of the dataset
- how estimates vary across isolates, years, fields, or treatments

The workflow shown here provides a starting point for both teaching and applied analyses in plant pathology.
